M\"obius Inversion: From Posets to Categories

### 2017-04-13 Drafts/Draft.tex
submitted version

\begin{abstract} M\"obius inversion is a general approach for reversing certain kinds of sums. We start by defining M\"obius inversion in its usual form, over a partially ordered set. We explore traditional applications in the integer divisibility lattice, including Euler's Totient. We then use the lattice of subgroups to count orbits under a group action. In the last two sections, we explore Tom Leinster's generalization of M\"obius inversion to categories and enriched categories, and investigate a data clustering algorithm.
\end{abstract}

@@

\begin{defn}
A \emph{partially ordered set}, often called a \emph{poset}, is a set $X$ together with a binary relation $\leq$, such that, for all $a,b,c \in X$:
\begin{itemize}
\item $a \leq a$ (reflexive)
\item If $a \leq b$ and $b \leq a$ then $a = b$ (antisymmetric)
\item If $a \leq b$ and $b \leq c$ then $a \leq c$ (transitive).
\end{itemize}
\end{defn}
Note that, for completeness, a poset should be labeled with both its elements and its relations - e.g. $(X, \leq)$. But often we just write $X$ when the relation is implied.
\begin{rmk}
Given $x_1, x_2$ in a poset $X$, we write $[x_1, x_2]$ to denote the set 
\[
\{y \in X: x_1 \leq y \leq x_2\}
\]
Replacing one of the square brackets with a parenthesis means we are omitting that element. So
\begin{align*}
[x_1, x_2) &= \{y \in X: x_1 \leq y < x_2\},\\
(x_1, x_2] &= \{y \in X: x_1 < y \leq x_2\}.
\end{align*}
Note that if $x_1 \nleq x_2, [x_1, x_2] = [x_1, x_2) = (x_1, x_2]= \emptyset.$
\end{rmk}
\begin{defn}
A poset $X$ is \emph{lower finite} if, for any $a \in x$, the set $\{ c \in X: c \leq a\}$ is finite.
\end{defn}
This condition will be necessary for many of the following facts and theorems, to ensure that certain sums have a finite number of nonzero terms. Note that in a lower finite poset, $[x_1, x_2]$ is finite for any $x_1, x_2$.

@@

The matrices and vectors we construct in this way might be infinite, so we have to be careful. In this chapter, all possibly infinite matrices we consider will be upper triangular (it is clear from the definition of $\zeta$ that this is the case here). Each term of the product of two upper triangular matrices is can be calculated from a finite number of matrix elements, even if the factor matrices are infinite. The same is true when an upper triangular matrix is left-multiplied by a row vector. Finally, because our matrices will be upper triangular and have ones on the diagonal, they will automatically be invertible (see Lemma \ref{inv_ex}), and can be inverted even if infinite. In subsequent chapters, we will restrict our attention to M\"obius inversion which uses only finite matrices.

@@

We now have two equivalent definitions of the M\"obius function, one using matrices (or linear operators) and vectors, and one directly considering elements and sums. We will continue to move back and forth between these two approaches throughout this investigation. This is because although the two definitions are completely equivalent, they each have advantages. Working with sums and elements helps us see the details of what is going on, and reminds us of the structure of our poset. It is also often the perspective from which motivating problems arise. On the other hand, using matrices and vectors greatly simplifies the equations we have to deal with, makes the manipulations themselves easier, and allows us to keep the bigger picture - the idea of inversion - in mind more easily. 
\section{Integer Lattices}
We now turn to a specific class of partially ordered sets, which will come up several times going forward.
\begin{defn}\label{lattice}
For any positive integer $n$, the \emph{integer lattice of dimension $n$}, which we denote $\Nn^n$, is the set of ordered $n$-tuples of non-negative integers. For any two elements $(a_1,\dots,a_n), (b_1,\dots,b_n) \in \Nn^n$, we say that $(a_1,\dots,a_n)\leq (b_1,\dots,b_n)$ if $a_i \leq b_i$ for each $i = 1,\cdots,n$.
\end{defn}
\begin{lemma}
$(\Nn^n, \leq)$ is a lower finite poset for any positive integer $n$. 
\end{lemma}
\begin{proof}
The poset axioms are easily checked. We also observe that for any $x \in (\Nn^n, \leq)$,
\[
\#\{y \in (\Nn^n, \leq):y \leq x\} = \prod_{i =1}^n x_i.
\]

@@

In this case, if $\leq$ represents the poset relation from Definition \ref{lattice}, then $m \leq n$ if, for each prime $p$, $m$'s factorization contains no more copies of $p$ than $n$'s factorization. This is the same as saying that $m|n$. So the poset relation becomes divisibility. We will explore this example more in the next sections.
\end{examp}
It turns out that several important posets have this lattice structure. So the goal is to determine the M\"obius function on certain subsets of the integer lattice, and then transfer those results to other partially ordered sets. To do so, we first need a lemma which applies to the M\"obius function on any poset.
\begin{lemma}\label{recurse}
Let $X$ be a lower finite partially ordered set, and let $\zeta, \mu$ be defined as usual. Then for all $a, b \in X$,
\[
\mu(a,b) = 
\begin{cases}
0 & a \nleq b\\
\delta(a,b)-\sum_{c \in [a,b)}\mu(a,c) & else.
\end{cases}
\]
\end{lemma}
\begin{rmk}
Note that the piecewise form is actually not necessary, because if $a \nleq b$ then the sum will be empty. So it is correct - albeit less clear- to just write
\[
\mu(a,b) = \delta(a,b)-\sum_{c \in [a,b)}\mu(a,c).
\]
\end{rmk}
\begin{proof}
We fix $a,b$, and then rearrange terms starting from Remark \ref{sum_form}:
\begin{align*}
\sum_{c \in X}\mu(a,c) \zeta(c,b) &= \delta(a,b)\\
\sum_{c \in [a,b)} \mu(a,c) + \mu(a,b) &= \delta(a,b)\\
\mu(a,b) &= \delta(a,b) - \sum_{c \in [a,b)} \mu(a,c) 
\end{align*}
\end{proof}
We are now ready to write down $\mu$ on the integer lattice.
\begin{thm}\label{int_lattice}
Let $L$ be the $n$-dimensional integer lattice, with the usual poset relation. Let $\zeta, \mu$ be defined as usual on $L$. Let $x=(x_1,\dots,x_n),y = (y_1,\dots,y_n)$ be elements of $L$. Then:
\[
\mu(x,y) =
\begin{cases}
0 & x\nleq y\\
0 & \exists i \in \Nn: y_i-x_i \geq 2\\
(-1)^{\sum_{i=1}^ny_i-x_i} & \text{otherwise}
\end{cases}
\]
\end{thm}

@@

Finally, suppose that for all $i = 1,\dots,n, y_i-x_i \leq 1$. Because $\sum_{i=1}^ny_i-x_i =k$, there must be $k$ indices where $x$ and $y$ differ by one, while at the other $n-k$ indices they are equal. So for each $j = 0,\dots,k-1$, there are $k \choose j$ elements $z$ of $X$ where $x \leq z < y$ and $\sum_{i = 1}^n y_i-z_i = j$. For each of these, by the inductive hypothesis, $\mu(x,z) = (-1)^j$. So, using Remark \ref{sum_form}, we can write
\begin{align*}
\mu(x,y)& = -\sum_{j = 1}^k {k \choose j} (-1)^j\\
&= (-1)^k.
\end{align*}
\end{proof}
This result will be powerful for determining values of $\mu$ on various posets whose structure is similar to an integer lattice. 
\begin{thm}\label{mapping}
Let $(X, \leq_X),(Y, \leq_Y)$ be lower finite posets. Let $x_1 \leq x_2 \in S$ be given. Suppose that there exists a bijective map
\[
\phi: [x_1, x_2] \leftrightarrow [\phi(x_1), \phi(x_2)]
\]
such that for all $z_1, z_2 \in [x_1, x_2]$,
\[ 
z_1 \leq_X z_2 \iff \phi(z_1) \leq_Y \phi(z_2).
\] 
Let $\mu_X, \mu_Y$ denote the M\"obius functions on $X$ and $Y$ respectively. Then:
\[
\mu_X(x_1, x_2) = \mu_Y(\phi(x_1), \phi(x_2)).
\]
\end{thm}
This theorem tells us that if we can find a portion of a familiar poset which has identical structure to a portion of an unknown poset, we can transfer what we know about values of $\mu$ on that portion from one to the other.
\begin{proof}
We induct on $\#[x_1, x_2]$. Because $x_1 \leq x_2$, if $[x_1, x_2] = 1$, then $x_1 = x_2$. So by Lemma \ref{recurse}, $\mu_X(x_1, x_2) = 1$. Because $\phi(x_1) = \phi(x_2), \mu_Y(\phi(x_1), \phi(x)2) = 1$ also.

Suppose the theorem is true if $\#[x_1, x_2] \leq k-1$ for some integer $k$. Fix $x_1, x_2$ such that $\#[x_1, x_2] = k$. Then by Lemma \ref{recurse},
\[
\mu_Y(\phi(x_1), \phi(x_2)) = -\sum_{y \in [\phi(x_1), \phi(x_2)]}\mu_Y(\phi(x_1), y)
\]
Because $\phi$ is an isomorphism which preserves $\leq$, we can rewrite this sum:
\[
\mu_Y(\phi(x_1), \phi(x_2)) = -\sum_{x \in [x_1, x_2]}\mu_Y(\phi(x_1), \phi(x))
\]
We can restrict $\phi$ to $[\phi(x_1), \phi(x)]$ to satisfy the hypotheses for all intermediate elements $x$. So we can apply the inductive hypothesis to all the terms of this sum:
\begin{align*}
\mu_Y(\phi(x_1), \phi(x_2)) &= -\sum_{x \in [x_1, x_2]}\mu_X(x_1, x)\\
&= \mu_X(x_1, x_2)
\end{align*}
\end{proof}
The following corollary follows immediately if we let $\phi$ be the inclusion map:
\begin{cor}\label{subset}
Let $X$ be a poset. Let $I$ be an interval in $X$. In other words, let $I = [z_1, z_2]$ for some $z_1, z_2 \in X$. Then
\[
\mu_I = \mu_X|_I.
\]
\end{cor}
In the next section, we explore a poset where this fact will be extremely useful.

@@

So $\phi:[d,n] \leftrightarrow [\phi(d), \phi(n)]$ is a bijection.

This allows us to use Theorem \ref{mapping}, and the result immediately follows.
\end{proof}

@@

\begin{figure}[h]
\[
\includegraphics{../Figures/s3_lattice.pdf}
\]
\caption{The lattice of subgroups of $S_3$.}
\end{figure}
So, for example, $\langle (1,2) \rangle \leq \langle e \rangle$, because $\langle (1,2) \rangle \supseteq \langle e \rangle$. 
\end{examp}
\begin{rmk}
We are ordering the poset this way for reasons that will soon become clear. But to reduce confusion, we will continue to use inclusion notation to define $\zeta$, and so $\mu$, rather than using $\leq$. I will be using $\subseteq$ rather than $\leq$ to denote subgroups, also to reduce confusion with the poset relation.
\end{rmk}
This puts us in the position to apply M\"obius inversion. We first define the usual functions on $\sub(G) \times \sub(G)$:
\begin{itemize}
\item Given a group $G$, let $\zeta:\sub(G) \times \sub(G) \rightarrow \Rr$ be defined by
\[
\zeta(a,b) = \begin{cases}
1 & A \supseteq B\\
0 & else
\end{cases}
\]
\item Let $\mu = \zeta^{-1}$, which exists by Lemma \ref{inv_ex}.
\end{itemize}
Recall that we manipulate $\zeta, \mu$ as matrices.

Now, Theorem \ref{mob_inv} tells us that, for any row vectors $f,g$ of length $|\sub{G}|$, 
\[
f = g\zeta \iff g = f\mu.
\]
Now, of course, the challenge is to determine how to formulate the problem such that inverting $\zeta$ will be useful.
\section{Counting Orbits}
Throughout this section, let $G$ be a finite group acting on a set $\Omega$.

@@

Now suppose that we are counting orbits by checking elements of $\Omega$ one at a time. Each time we see an element $x \in \Omega$, we ask it for the size of its orbit, and then add $\frac 1 {|\orb_G(x)|}$ to a running total. Let $O$ be a single orbit of $\Omega$ under $G$ where $|O|=n$. Then we will see $n$ elements of $\Omega$ which belong to $O$, and for each of them, we will add $\frac 1n$ to our running total. This means we add exactly 1 for $O$. The same is true of all the orbits in $\Omega$. So our final total will be the size of $G/X$.

@@

\begin{equation}\label{eq_counting}
|G/X| = \sum_{x \in \Omega}\frac 1 {|\orb_G(x)|}.
\end{equation}

@@

\[|G| = |G(x)||\stab(x, G)|.\]

Because $G$ is finite, this implies that \[|G(x)| = \frac{|G|}{|\stab(x, G)|}.\]

Starting from a trivial fact, we have:
\begin{align*}
|\orb(\Omega, G)| &= \sum_{O \in \orb(\Omega, G)}1\\
&= \sum_{O \in \orb(\Omega, G)}\left(\sum_{x \in O}\frac{1}{|O|}\right)\\
&= \sum_{x \in \Omega}\frac{1}{|G(x)|}
\end{align*}

Substituting, we now have the relationship:
\[
|\orb(\Omega, G)| = \frac{1}{|G|}\sum_{x \in \Omega}|\stab(x, G)|
\]
Each element of $\Omega$ has exactly one stabilizer in $G$. This means that we can partition the above sum as follows:
\[
|\orb(\Omega, G)| = \frac{1}{|G|}\sum_{A \in \sub(G)}\left(\sum_{x \in \Omega: \stab(x) = A}|A|\right)
\]
Using the definitions of $g$ and $\sigma$, we can rewrite to get the relationship we need:
\begin{align*}
|\orb(\Omega, G)| &= \frac{1}{|G|}\sum_{A \in \sub(G)}g(A)|A|\\
&= \frac {1}{|G|}g \cdot \sigma
\end{align*}

On to Step 2. It is clear that $f_A = \sum_{B \supseteq A} g_A$. Another way of putting that is that $f = g\zeta$. As usual, this implies that $g = f\mu$. So M\"obius inversion makes this step trivial. Putting it all together:

\[|\orb(\Omega, G)| = \frac {1}{|G|}f\mu \cdot \sigma.\]

Recall that $\sigma$ and $\mu$ do not depend on the details of the group action, or on $\Omega$ at all. They only depend on $G$. So it is useful to rearrange this expression to isolate $f$, the only part which does depend on the action. This gives us the central result of this chapter:
\begin{thm}\label{orbit_count}
Let $G$ act on a set $\Omega$. Let $f, \mu, \sigma$ be defined as above. Then
\[
|\orb(\Omega, G)| = \frac {1}{|G|}(\mu \cdot \sigma)^Tf^T
\]
\end{thm}

@@

This implies that
\[
\mu = \zeta^{-1} = \begin{bmatrix} 1 & -1 & 0\\ 0 & 1 & -1 \\ 0 & 0 & 1 \end{bmatrix}.
\]
Next, we find $\sigma$. Recall that $\sigma$ is the vector of subgroup sizes. So 
\[
\sigma = [4, 2, 1].
\]
Finally, note that $|C_4|=4$. So
\begin{align*}
\lambda &= \frac 14 \mu \sigma^T\\
&= \frac 14 
\begin{bmatrix} 1 & -1 & 0\\ 0 & 1 & -1 \\ 0 & 0 & 1 \end{bmatrix}
\begin{bmatrix} 4 \\ 2 \\ 1 \end{bmatrix}\\
&= \frac 14 \begin{bmatrix} 2 \\ 1 \\ 1\end{bmatrix}
\end{align*}
We determined $\lambda$ before even deciding what set $G$ might act on. So all the work we have done so far can be reused for a variety of actions.

@@

We can first solve this problem by inspection, or by "brute force," by just enumerating the possibilities. This way, when we solve it with $\lambda$, we can determine whether we found the right answer.
\begin{description}
\item[Zero rooks:] There is one option with zero rooks: The empty board.
\[
\includegraphics{../Figures/zero_rooks.pdf}
\]
\item[One rook:] We can put one rook on any of the four squares. However, we can rotate the board to obtain any of these from the others. So there is just one distinct configuration. \\
\[
\makebox[\textwidth][c]{\includegraphics{../Figures/one_rook.pdf}}
\]
\item[Two rooks:] We can put rooks on opposite corners. There are two such pairs, but we can rotate the board and obtain either pair from the other. So, again, there is only one distinct possibility.
\[
\makebox[\textwidth][c]{\includegraphics{../Figures/two_rooks.pdf}}
\]
\end{description}
So there are a total of three distinct arrangements of rooks. Now we can try to solve this problem using $\lambda$.

@@

Luckily, we know how to count orbits. All we have to do is find $f$. We go through the subgroups of $C_4$.
\begin{itemize}
\item $C_4$ only fixes the empty board. Any other arrangement can be changed by a rotation. So $f(1) = 1$.
\item The subgroup $\langle \alpha^2 \rangle$ contains the identity element as well as the $180^\circ$ rotation. This subgroup fixes both arrangements with two rooks, as well as the empty board. So $f(2) = 3$.
\item The subgroup $\langle e \rangle$ fixes all the arrangements. There are 2 ways to place two rooks, 4 ways to place one rook, and 1 way to place zero rooks. So $f(2) = 7$.
\end{itemize}
Then all we have to do is compute $\lambda f^T$.
\begin{align*}
\lambda f^T &= \frac 14 [2, 1, 1] \begin{bmatrix} 1 \\ 3 \\ 7\end{bmatrix} \\
&= \frac 14 12\\
&=3.
\end{align*}
So we indeed get the correct answer. 
\end{examp}

In this example, it's not clear that using $\lambda$ actually saved work - in fact, it seemed to take longer. But if we were to increase $n$, we could still determine the number of distinct arrangements relatively easily, because $\lambda$ wouldn't change. Furthermore, in many situations (including this one) it is much easier to determine how many \emph{total} arrangements are fixed by a subgroup that it is to count the number of \emph{distinct} arrangements. We can look at another example, which is more difficult to solve by enumeration. 
\begin{examp}[Example \ref{ex:coloring_init}, cont]\label{ex:coloring}
We now have the tools to solve the problem from Example \ref{ex:coloring_init}. We briefly restate the setup: Suppose we have a $4 \times 4$ board, and we would like to color each square one of $k$ colors. We would like to consider colorings identical if one can be obtained from the other by rotating the board. The goal is to count the number of distinct colorings.

>>>>

\begin{abstract} M\"obius inversion is a general approach for reversing certain kinds of sums. We start by defining M\"obius inversion in its usual form, over a partially ordered set. We explore traditional applications in the integer divisibility lattice, including Euler's Totient. We then use the lattice of subgroups to count orbits under a group action, and use the lattice of subsets to prove the Inclusion-Exclusion Principle. In the last two sections, we explore Tom Leinster's generalization of M\"obius inversion to categories and enriched categories, and investigate a data clustering algorithm.
\end{abstract}

@@

\begin{defn}
A \emph{partially ordered set}, often called a \emph{poset}, is a set $X$ together with a binary relation $\leq$, such that, for all $a,b,c \in X$:
\begin{itemize}
\item $a \leq a$ (reflexive)
\item If $a \leq b$ and $b \leq a$ then $a = b$ (antisymmetric)
\item If $a \leq b$ and $b \leq c$ then $a \leq c$ (transitive).
\end{itemize}
\end{defn}
Note that, for completeness, a poset should be labeled with both its elements and its relations - e.g. $(X, \leq)$. But often we just write $X$ when the relation is implied.
\begin{rmk}
Given $x_1, x_2$ in a poset $X$, we write $[x_1, x_2]$ to denote the set 
\[
\{y \in X: x_1 \leq y \leq x_2\}
\]
Replacing one of the square brackets with a parenthesis means we are omitting the bound on that end. So
\begin{align*}
[x_1, x_2) &= \{y \in X: x_1 \leq y < x_2\},\\
(x_1, x_2] &= \{y \in X: x_1 < y \leq x_2\}.
\end{align*}
Note that if $x_1 \nleq x_2, [x_1, x_2] = [x_1, x_2) = (x_1, x_2]= \emptyset.$
\end{rmk}
\begin{defn}
A poset $X$ is \emph{lower finite} if, for any $a \in x$, the set $\{ c \in X: c \leq a\}$ is finite.
\end{defn}
This condition will be necessary for many of the following facts and theorems, to ensure that certain sums have a finite number of nonzero terms. Note that in a lower finite poset, $[x_1, x_2]$ is finite for any $x_1, x_2$.

@@

The matrices and vectors we construct in this way might be infinite, so we have to be careful. In this chapter, all possibly infinite matrices we consider will be upper triangular (it is clear from the definition of $\zeta$ that this is the case here). Each term of the product of two upper triangular matrices is can be calculated from a finite number of matrix elements, even if the factor matrices are infinite. The same is true when an upper triangular matrix is left-multiplied by a row vector. Finally, because our matrices will be upper triangular and have ones on the diagonal, they will automatically be invertible (see Lemma \ref{inv_ex}), have upper triangular inverses, and can be inverted even if infinite. In subsequent chapters, we will restrict our attention to M\"obius inversion which uses only finite matrices.

@@

We now have two equivalent definitions of the M\"obius function, one using matrices (or linear operators) and vectors, and one directly considering elements and sums. We will continue to move back and forth between these two approaches throughout this investigation. This is because although the two definitions are completely equivalent, they each have advantages. Working with sums and elements helps us see the details of what is going on, and reminds us of the structure of our poset. It is also often the perspective from which motivating problems arise. On the other hand, using matrices and vectors greatly simplifies the equations we have to deal with, makes the manipulations themselves easier, and allows us to keep the bigger picture - the idea of inversion - in mind more easily. 
\section{Integer Lattices}
We now turn to a specific class of partially ordered sets, which will come up several times going forward.
\begin{defn}\label{lattice}
For any positive integer $n$, the \emph{integer lattice of dimension $n$}, which we denote $\Nn^n$, is the set of ordered $n$-tuples of non-negative integers. For any two elements $(a_1,\dots,a_n), (b_1,\dots,b_n) \in \Nn^n$, we say that $(a_1,\dots,a_n)\leq (b_1,\dots,b_n)$ if $a_i \leq b_i$ for each $i = 1,\cdots,n$.
\end{defn}
\begin{lemma}
$(\Nn^n, \leq)$ is a lower finite poset for any positive integer $n$. 
\end{lemma}
\begin{proof}
The poset axioms are easily checked. To see why the poset is lower finite, we observe that for any $x \in (\Nn^n, \leq)$,
\[
\#\{y \in (\Nn^n, \leq):y \leq x\} = \prod_{i =1}^n x_i.
\]

@@

In this case, if $\leq$ represents the poset relation from Definition \ref{lattice}, then $m \leq n$ if, for each prime $p$, $m$'s factorization contains no more copies of $p$ than $n$'s factorization. This is the same as saying that $m|n$. So the poset relation becomes divisibility. We will explore this example more in the next sections.
\end{examp}
It turns out that several important posets have this lattice structure. So the goal is to determine the M\"obius function on certain subsets of the integer lattice, and then transfer those results to other partially ordered sets. To do so, we first need a lemma which applies to the M\"obius function on any poset.
\begin{lemma}\label{recurse}
Let $X$ be a lower finite partially ordered set, and let $\zeta, \mu$ be defined as usual. Then for all $a, b \in X$,
\[
\mu(a,b) = 
\begin{cases}
0 & a \nleq b\\
\delta(a,b)-\sum_{c \in [a,b)}\mu(a,c) & else.
\end{cases}
\]
\end{lemma}
\begin{rmk}
Note that the piecewise form is actually not necessary, because if $a \nleq b$ then the sum will be empty. So it is also correct to just write
\[
\mu(a,b) = \delta(a,b)-\sum_{c \in [a,b)}\mu(a,c).
\]
\end{rmk}
\begin{proof}
We fix $a,b$, and then rearrange terms starting from Remark \ref{sum_form}:
\begin{align*}
\sum_{c \in X}\mu(a,c) \zeta(c,b) &= \delta(a,b)\\
\sum_{c \in [a,b)} \mu(a,c) + \mu(a,b) &= \delta(a,b)\\
\mu(a,b) &= \delta(a,b) - \sum_{c \in [a,b)} \mu(a,c) 
\end{align*}
\end{proof}
We are now ready to write down $\mu$ on the integer lattice.
\begin{thm}\label{int_lattice}
Let $L$ be the $n$-dimensional integer lattice, with the usual poset relation. Let $\zeta, \mu$ be defined as usual on $L$. Let $x=(x_1,\dots,x_n),y = (y_1,\dots,y_n)$ be elements of $L$. Then:
\[
\mu(x,y) =
\begin{cases}
0 & x\nleq y\\
0 & \exists i \in \Nn: y_i-x_i \geq 2\\
(-1)^{\sum_{i=1}^ny_i-x_i} & \text{otherwise}
\end{cases}
\]
\end{thm}

@@

Finally, suppose that for all $i = 1,\dots,n, y_i-x_i \leq 1$. Because $\sum_{i=1}^ny_i-x_i =k$, there must be $k$ indices where $x$ and $y$ differ by one, while at the other $n-k$ indices they are equal. So for each $j = 0,\dots,k-1$, there are $k \choose j$ elements $z$ of $X$ where $x \leq z < y$ and $\sum_{i = 1}^n y_i-z_i = j$. For each of these, by the inductive hypothesis, $\mu(x,z) = (-1)^j$. So, using Remark \ref{sum_form}, we can write
\begin{align*}
\mu(x,y)& = -\sum_{j = 1}^k {k \choose j} (-1)^j\\
&= (-1)^k.
\end{align*}
\end{proof}
This result will be powerful for determining values of $\mu$ on various posets whose structure is similar to an integer lattice. 
\begin{thm}\label{mapping}
Let $(X, \leq_X),(Y, \leq_Y)$ be lower finite posets. Let $x_1 \leq x_2 \in S$ be given. Suppose that there exists a bijective map
\[
\phi: [x_1, x_2] \leftrightarrow [\phi(x_1), \phi(x_2)]
\]
such that for all $z_1, z_2 \in [x_1, x_2]$,
\[ 
z_1 \leq_X z_2 \iff \phi(z_1) \leq_Y \phi(z_2).
\] 
Let $\mu_X, \mu_Y$ denote the M\"obius functions on $X$ and $Y$ respectively. Then:
\[
\mu_X(x_1, x_2) = \mu_Y(\phi(x_1), \phi(x_2)).
\]
\end{thm}
This theorem tells us that if we can find a portion of a familiar poset which has identical structure to a portion of an unknown poset, we can transfer what we know about values of $\mu$ on that portion from one to the other.
\begin{proof}
We induct on $\#[x_1, x_2]$, starting with the base case. Because $x_1 \leq x_2$, if $[x_1, x_2] = 1$, then $x_1 = x_2$. So by Lemma \ref{recurse}, $\mu_X(x_1, x_2) = 1$. Because $\phi(x_1) = \phi(x_2)$, we know that $\mu_Y(\phi(x_1), \phi(x)2) = 1$. So the M\"obius functions take the same value.

Suppose the theorem is true if $\#[x_1, x_2] \leq k-1$ for some integer $k$. Fix $x_1, x_2$ such that $\#[x_1, x_2] = k$. Then by Lemma \ref{recurse},
\[
\mu_Y(\phi(x_1), \phi(x_2)) = -\sum_{y \in [\phi(x_1), \phi(x_2))}\mu_Y(\phi(x_1), y)
\]
Because $\phi$ is an isomorphism which preserves $\leq$, we can rewrite this sum:
\[
\mu_Y(\phi(x_1), \phi(x_2)) = -\sum_{x \in [x_1, x_2)}\mu_Y(\phi(x_1), \phi(x))
\]
We can restrict $\phi$ to $[x_1, x]$ to satisfy the hypotheses for all intermediate elements $x$. So we can apply the inductive hypothesis to all the terms of this sum:
\begin{align*}
\mu_Y(\phi(x_1), \phi(x_2)) &= -\sum_{x \in [x_1, x_2)}\mu_X(x_1, x)\\
&= \mu_X(x_1, x_2)
\end{align*}
\end{proof}
The following corollary follows immediately if we let $\phi$ be the inclusion map:
\begin{cor}\label{subset}
Let $X$ be a poset. Let $I$ be an interval in $X$. In other words, let $I = [z_1, z_2]$ for some $z_1, z_2 \in X$. Then
\[
\mu_I = \mu_X|_I.
\]
\end{cor}
In the next section, we explore a poset where this fact will be extremely useful.

@@

So $\phi:[d,n] \leftrightarrow [\phi(d), \phi(n)]$ is a bijection. This allows us to use Theorem \ref{mapping}, and the result immediately follows.
\end{proof}

@@

\begin{figure}[h]
\[
\includegraphics{../Figures/s3_lattice.pdf}
\]
\caption{The lattice of subgroups of $S_3$.}
\end{figure}
So, for example, $\langle (1,2) \rangle \leq \langle e \rangle$, because $\langle (1,2) \rangle \supseteq \langle e \rangle$. 
\end{examp}

This puts us in the position to apply M\"obius inversion. We first define the usual functions on $\sub(G) \times \sub(G)$:
\begin{itemize}
\item Given a group $G$, let $\zeta:\sub(G) \times \sub(G) \rightarrow \Rr$ be defined by
\[
\zeta(a,b) = \begin{cases}
1 & A \supseteq B\\
0 & else
\end{cases}
\]
\item Let $\mu = \zeta^{-1}$, which exists by Lemma \ref{inv_ex}.
\end{itemize}
Recall that we manipulate $\zeta, \mu$ as matrices.

Theorem \ref{mob_inv} tells us that, for any row vectors $f,g$ of length $|\sub{G}|$, 
\[
f = g\zeta \iff g = f\mu.
\]
Now, of course, the challenge is to determine how to formulate the problem of counting orbits such that inverting $\zeta$ will be useful.
\section{Counting Orbits}
Throughout this section, let $G$ be a finite group acting on a finite set $\Omega$.

@@

Now suppose that we are counting orbits by checking elements of $\Omega$ one at a time. Each time we see an element $x \in \Omega$, we ask it for the size of its orbit, and then add $\frac 1 {|\orb_G(x)|}$ to a running total. Let $O$ be a single orbit of $\Omega$ under $G$ where $|O|=n$. Then we will see $n$ elements of $\Omega$ which belong to $O$, and for each of them, we will add $\frac 1n$ to our running total. This means we add exactly 1 for $O$. The same is true of all the orbits in $\Omega$. So our final total will be the size of $\Omega/G$.

@@

\begin{equation}\label{eq_counting}
|\Omega/G| = \sum_{x \in \Omega}\frac 1 {|\orb_G(x)|}.
\end{equation}

@@

\[|G| = |\orb_G(x)||\stab(x, G)|.\]

Because $G$ is finite, this implies that \begin{equation}|\orb_G(x)| = \frac{|G|}{|\stab(x, G)|}.\end{equation} \label{eq:orb_stab}

Starting from a trivial fact, we can recover \eqref{eq_counting}:
\begin{align*}
|\Omega/G| &= \sum_{O \in \Omega/G}1\\
&= \sum_{O \in \Omega/G}\left(\sum_{x \in O}\frac{1}{|O|}\right)\\
&= \sum_{x \in \Omega}\frac{1}{|\orb_G(x)|}
\end{align*}

Substituting \eqref{eq:orb_stab}, we now have the relationship:
\[
|\Omega/G| = \frac{1}{|G|}\sum_{x \in \Omega}|\stab(x, G)|
\]
Each element of $\Omega$ has exactly one stabilizer in $G$. This means that we can partition the above sum as follows:
\[
|\Omega/G| = \frac{1}{|G|}\sum_{A \in \sub(G)}\left(\sum_{x \in \Omega: \stab(x) = A}|A|\right)
\]
Using the definitions of $g$ and $\sigma$, we can rewrite to get the relationship we need:
\begin{align*}
|\Omega/G| &= \frac{1}{|G|}\sum_{A \in \sub(G)}g(A)|A|\\
&= \frac {1}{|G|}g \cdot \sigma
\end{align*}

On to Step 2. It is clear from our definitions that \[f_A = \sum_{B \supseteq A} g_A.\] Another way of putting it is that $f = g\zeta$. As usual, this implies that $g = f\mu$. So M\"obius inversion makes this step trivial. Putting it all together:

\[|\Omega/G| = \frac {1}{|G|}f\mu \cdot \sigma.\]

Recall that $\sigma$ and $\mu$ do not depend on the details of the group action, or on $\Omega$ at all. They only depend on $G$. So it is useful to rearrange this expression to isolate $f$, the only part which does depend on the action. This gives us the central result of this chapter:
\begin{thm}\label{orbit_count}
Let $G$ act on a set $\Omega$. Let $f, \mu, \sigma$ be defined as above. Then
\[
|\Omega/G| = \frac {1}{|G|}(\mu \cdot \sigma)^Tf^T
\]
\end{thm}

@@

This implies that
\[
\mu = \zeta^{-1} = \begin{bmatrix} 1 & -1 & 0\\ 0 & 1 & -1 \\ 0 & 0 & 1 \end{bmatrix}.
\]
Next, we find $\sigma$. Recall that $\sigma$ is the vector of subgroup sizes. So 
\[
\sigma = [4, 2, 1].
\]
Finally, note that $|C_4|=4$. So
\begin{align*}
\lambda &= \frac 14 \mu \sigma^T\\
&= \frac 14 
\begin{bmatrix} 1 & -1 & 0\\ 0 & 1 & -1 \\ 0 & 0 & 1 \end{bmatrix}
\begin{bmatrix} 4 \\ 2 \\ 1 \end{bmatrix}\\
&= \frac 14 \begin{bmatrix} 2 \\ 1 \\ 1\end{bmatrix}.
\end{align*}
We determined $\lambda$ before even deciding what set $G$ might act on. So all the work we have done so far can be reused for a variety of actions.

@@

We can first solve this problem by inspection, or by ``brute force,'' by just enumerating the possibilities. This way, when we solve it with $\lambda$, we can determine whether we found the right answer.
\begin{description}
\item[Zero rooks:] There is one option with zero rooks: The empty board.
\[
\includegraphics{../Figures/zero_rooks.pdf}
\]
\item[One rook:] We can put one rook on any of the four squares. However, we can rotate the board to obtain any of these from the others. So there is just one distinct configuration. \\
\[
\makebox[\textwidth][c]{\includegraphics{../Figures/one_rook.pdf}}
\]
\item[Two rooks:] We can put rooks on opposite corners. There are two such pairs, but we can rotate the board and obtain either pair from the other. So, again, there is only one distinct possibility.
\[
\makebox[\textwidth][c]{\includegraphics{../Figures/two_rooks.pdf}}
\]
\end{description}
So there are a total of three distinct arrangements of rooks. Now we can try to solve this problem using $\lambda$.

@@

Luckily, we know how to count orbits. All we have to do is find $f$. We go through the subgroups of $C_4$.
\begin{itemize}
\item $C_4$ only fixes the empty board. Any other arrangement can be changed by a rotation. So $f(1) = 1$.
\item The subgroup $\langle \alpha^2 \rangle$ contains the identity element as well as the $180^\circ$ rotation. This subgroup fixes both arrangements with two rooks, as well as the empty board. So $f(2) = 3$.
\item The subgroup $\langle e \rangle$ fixes all the arrangements. There are 2 ways to place two rooks, 4 ways to place one rook, and 1 way to place zero rooks. So $f(2) = 7$.
\end{itemize}
Then all we have to do is compute $\lambda f^T$.
\begin{align*}
\lambda f^T &= \frac 14 [2, 1, 1] \begin{bmatrix} 1 \\ 3 \\ 7\end{bmatrix} \\
&= \frac 14\cdot 12\\
&=3.
\end{align*}
So we indeed get the correct answer. 
\end{examp}

In this example, it's not clear that using $\lambda$ actually saved work. But if we were to increase $n$, we could still determine the number of distinct arrangements relatively easily, because $\lambda$ wouldn't change. Furthermore, in many situations (including this one) it is much easier to determine how many \emph{total} arrangements are fixed by a subgroup that it is to count the number of \emph{distinct} arrangements. We can look at another example, which is more difficult to solve by enumeration. 
\begin{examp}[Example \ref{ex:coloring_init}, cont]\label{ex:coloring}
We now have the tools to solve the problem from Example \ref{ex:coloring_init}. We briefly restate the setup: Suppose we have a $4 \times 4$ board, and we would like to color each square one of $k$ colors. We would like to consider colorings identical if one can be obtained from the other by rotating the board. The goal is to count the number of distinct colorings.
